Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-03-28 ms/ms.Rnw
add figs, tweak

%\bibpunct{(}{)}{;}{a}{}{,}
\usepackage{natbib}
\usepackage{palatino}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames]{xcolor} %% for non-standard cite color
\usepackage{graphicx}
\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
\usepackage{url}
\usepackage{enumerate}
\usepackage[left=50pt,
            textwidth=345pt,
            marginparsep=25pt,
            marginparwidth=124pt]{geometry}
%\usepackage[margin=1in]{geometry}

@@

<<setup,echo=FALSE>>=
library("knitr")
opts_chunk$set(echo=FALSE,error=FALSE,fig.width=8,fig.height=4) #
@

<<pkgs,warning=FALSE,message=FALSE>>=
library("ggplot2"); theme_set(theme_classic())
library("MASS")
scale_colour_discrete <- function(...,palette="Set1") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Set1") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("knitr")
## devtools::install_github("lionel-/ggstance")
library("ggstance")
@

@@

\todo{define each summary statistic and explain what it tells us}
\begin{description}
\item[peak time] ...
\item[peak virulence] ...
\item[relative peak virulence] ...
\item[equilibrium virulence] ...
\end{description}

@@

Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a). Our model produces lower peak virulence ($\approx ?$ vs. $\approx ?$) and equilibrium virulence $\approx ?$ vs. $\approx?$) than Shirreff's, probably because \ldots \todo{describe more; is this because we 
can't match both R0 and r? Can we match r exactly?}. (We're matching epidemic curve, not virulence curve!)

  Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b). Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence.

 For the baseline set of parameters, with the initial epidemic spread rate scaled to be the
 \todo{just double-checking: Figure 1 shows results from the implicit model, right?}

 Throughout Fig 1 we're using Shirreff's parameters (Beta1 = 2.76, Duration1 = 0.25, Beta3 = 0.76, Duration3 = 0.75, cmeanbase = 1.25, alphaDist = c(min = 2, max = 7,  delta = 0.1))

@@

<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",dev="tikz",fig.width=6,cache=TRUE>>=
gg_virtraj <- ggplot(sum_list[["vir_mat"]],
       aes(tvec,mean,ymin=lwr,ymax=upr))+
    geom_line(aes(colour=model,linetype=model),lwd=1.5)+
    geom_ribbon(aes(fill=model),alpha=0.3)+
    labs(x="time",y=vlab)
print(gg_virtraj)
@

<<unidist,fig.cap="Univariate distributions of summary statistics. Note: point for eqvir/random is a placeholder, will have to be explained",fig.width=8,fig.height=4.8>>=
sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
mL <- melt(sL,id.vars=c("model","run"))
## horrible hack (but doesn't help); subsample
## w <- which(mL$model=="random")
## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
w <- with(mL,which(model=="random" & variable=="eq_vir"))
rval <- mean(mL$value[w])
mLw <- droplevels(mL[-w,])
ggplot(mLw,aes(value,model,fill=model))+
    geom_violinh(width=1)+
    facet_wrap(~variable,scale="free_x")+
    guides(fill=guide_legend(reverse=TRUE))+
    labs(y="")+
    geom_point(data=data.frame(model="random",variable="eq_vir",
                               value=rval),pch=22,size=3,show.legend=FALSE)
@

@@

>>>>

%\bibpunct{(}{)}{;}{a}{}{,}
\usepackage{natbib}
\usepackage{palatino}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{enumerate}
\usepackage[colorlinks = true, linkcolor = blue, urlcolor=blue,citecolor = MidnightBlue]{hyperref}
\usepackage[left=50pt,
            textwidth=345pt,
            marginparsep=25pt,
            marginparwidth=124pt]{geometry}
%\usepackage[margin=1in]{geometry}

@@

<<setup,echo=FALSE>>=
library("knitr")
opts_chunk$set(echo=FALSE,error=FALSE,fig.width=8,fig.height=4) #
## https://github.com/yihui/tikzDevice/issues/46 
local({
    hook_document <- knitr::knit_hooks$get('document')
    knitr::knit_hooks$set(document = function(x){
        res <- hook_document(x)
        gsub('\\usepackage[]{color}', '\\usepackage[usenames,dvipsnames]{xcolor}', res, fixed = TRUE)
  })
})
@

<<pkgs,warning=FALSE,message=FALSE>>=
library("ggplot2"); theme_set(theme_classic())
library("MASS")
## use Dark2 rather than Set1 because colour #6 of Set1 is yellow (ugh/too light)
scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
library("gridExtra")
library("plyr")  ## for ldply()
library("dplyr") ## for full_join()
library("reshape2")  ## for melt()
library("GGally")
library("knitr")
## devtools::install_github("lionel-/ggstance")
library("ggstance")
@

@@

@@

Our simplifications of \shir's model reproduce the qualitative behaviour reasonably well; as $r$ decreases from 0.084 to 0.42 (the latter value matching the initial rate of increase in prevalence in \shir's full model) the initial trajectory of increasing virulence brackets the rate from the original model (Figure~\ref{fig:panel3}a). However, our model produces lower peak virulence ($\approx ?$ vs. $\approx ?$) and equilibrium virulence $\approx ?$ vs. $\approx?$) than Shirreff's, even for matching initial incidence (i.e., $r=0.042$).

Changing the initial infectious density ($I(0)$), while it produces the expected changes in the initial epidemic trajectory (Supplementary material), has little effect on the virulence trajectory, making the virulence peaks slightly later and larger as $I(0)$ decreases and allows a longer epidemic phase before the transition to endemic dynamics (Figure~\ref{fig:panel3}b). Decreasing the initial virulence similarly but more strongly leads to progressively later, larger peaks in virulence (Figure~\ref{fig:panel3}c).

\todo{note that in Fig 1 we're using Shirreff's parameters}
\todo{add Shir. params to table 1: Beta1=2.76, Dur1 = 0.25, Beta3 = 0.76, Dur3 = 0.75, c = 1.25, alpha=\{2,7,0.1\}}

<<setup2,echo=FALSE>>=
orig_sum_labs <- c("peak_time","peak_vir","eq_vir","rel_peak")
new_sum_labs <- c("peak time","peak virulence","equil virulence","relative peak")
fixfac2 <- function(x) factor(x,
                       levels=orig_sum_labs,
                       labels=new_sum_labs)
@

@@

<<virtraj,fig.cap="Envelopes of virulence trajectories under all models",fig.width=6,cache=TRUE,dev="tikz">>=
gg_virtraj <- ggplot(sum_list[["vir_mat"]],
                     aes(tvec,mean,ymin=lwr,ymax=upr))+
    geom_line(aes(colour=model,linetype=model),lwd=1.5)+
    geom_ribbon(aes(fill=model),alpha=0.3)+
    labs(x="time",y=vlab)
print(gg_virtraj)
@

<<unidist,fig.cap="Univariate distributions of summary statistics. Note: point for eqvir/random is a placeholder, will have to be explained",fig.width=8,fig.height=4.8>>=
sL <- transform(sum_list[["sum_mat"]],rel_peak=peak_vir/eq_vir)
mL <- melt(sL,id.vars=c("model","run"))
## horrible hack (but doesn't help); subsample
## w <- which(mL$model=="random")
## mLw <- mL[-sample(w,size=length(w)*9/10,replace=FALSE),]
w <- with(mL,which(model=="random" & variable=="eq_vir"))
rval <- mean(mL$value[w])
mLw <- droplevels(mL[-w,])
mLw$variable <- fixfac2(mLw$variable)
ggplot(mLw,aes(value,model,fill=model))+
    geom_violinh(width=1)+
    facet_wrap(~variable,scale="free_x")+
    guides(fill=guide_legend(reverse=TRUE))+
    labs(y="")+
    geom_point(data=data.frame(model="random",variable="equil virulence",
                               value=rval),pch=22,size=3,show.legend=FALSE)
@

<<pairplot,echo=FALSE,fig.cap="pairs plot",fig.width=7,fig.height=7>>=
## names(sL)[na.omit(match(names(sL),orig_sum_labs))] <- paste0("`",new_sum_labs,"`")
ggp1 <- ggpairs(sL,
        mapping = ggplot2::aes(color = model),
        columns=3:6,
        lower = list(continuous = wrap("points", alpha = 0.5,size=0.5)),
        diag = list(continuous = "blankDiag"),
        upper = list(continuous = "blank"))+
    theme(panel.margin=grid::unit(0,"lines"))
## http://stackoverflow.com/questions/14711550/is-there-a-way-to-change-the-color-palette-for-ggallyggpairs-using-ggplot
## have to change plot one panel at a time
cbscales <- list(scale_colour_brewer(palette = 'Dark2'),
                 scale_fill_brewer(palette = 'Dark2'))  ## DRY
for (row in seq_len(ggp1$nrow))
    for (col in seq_len(ggp1$ncol))
        ggp1[row, col] <- ggp1[row, col] + cbscales[[(row==col)+1]]
## hack to remove top row/right column
ggp1$nrow <- ggp1$ncol <- 3
ggp1$plots <- ggp1$plots[c(5:7,9:11,13:15)]
print(ggp1)
@

@@

<<plot_sens,dev="png",echo=FALSE,fig.width=10,fig.height=6>>=
ggplot(mL2,aes(LHSval,sumval,colour=model))+
    geom_point(pch=".",alpha=0.5)+
    facet_grid(sumvar~LHSvar,scales="free")+
    geom_smooth(se=FALSE)+labs(x="",y="")+
    scale_x_log10()
@
